\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage[shortlabels]{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{fancyvrb}
\usepackage{parskip}
\usepackage{microtype}
\usepackage{tabularx}
\usepackage{authblk}
\usepackage{abstract}
\sloppy
\emergencystretch=3em
\hbadness=10000

% Theorem environments (academic-paper standard).
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{construction}[theorem]{Construction}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Role tag (kept for cross-reference with implementation notes).
\newcommand{\role}[1]{\textsc{\small[#1]}\,}

% Subsection-level phase header.
\newcommand{\phase}[2]{%
  \subsection[#1]{\texorpdfstring{\textbf{#1} --- #2}{#1 --- #2}}%
}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue!60!black,
  citecolor=blue!60!black,
  pdftitle={Ziren Multilinear PCS Pipeline: From WHIR Soundness Fixes to a BaseFold Production Backend},
  pdfauthor={ProjectZKM Research}
}

\definecolor{codebg}{gray}{0.95}
\definecolor{codeborder}{gray}{0.85}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{codeborder},
  framesep=4pt,
  xleftmargin=6pt,
  xrightmargin=6pt,
  breaklines=true,
  breakatwhitespace=false,
  postbreak=\mbox{\textcolor{gray}{$\hookrightarrow$}\space},
  columns=fullflexible,
  keepspaces=true,
  aboveskip=0.6em,
  belowskip=0.6em,
  showstringspaces=false,
}

\renewcommand{\abstractname}{\large Abstract}

\title{\Large\textbf{Ziren Multilinear PCS Pipeline}\\[0.3em]
  \large From WHIR Soundness Fixes to a BaseFold Production Backend%
  \\[0.5em]
  \normalsize with Jagged Packing, Late-Binding, and Per-Chip Commitments}

\author[1]{ProjectZKM Research}
\affil[1]{ProjectZKM}

\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
We document a two-phase upgrade of the Ziren zkVM's multilinear
polynomial commitment stack.  The first phase is a production-sound
integration of the WHIR multilinear PCS, closing three soundness
gaps in the upstream WHIR fast path: (i)~unbound transition
constraints, fixed by a sumcheck-based zerocheck; (ii)~unbound
lookup arguments, fixed by a sumcheck reduction of LogUp-GKR; and
(iii)~unbound per-row MLE openings, fixed by a \emph{late-binding}
WHIR adapter that allows evaluation points to be sampled
\emph{after} commitment.  A \emph{cross-binding} spot-check
construction ties the FRI commit and the WHIR late-binding commit
to the same trace, ruling out a divergent-trace attack.  A
\emph{jagged late-binding} reduction combines variable-height
chip-trace packing with the late-binding adapter, achieving
$57$--$85\times$ smaller per-shard proofs at $1.3$--$2.1\times$
prover cost.

The second phase replaces WHIR entirely with a
\emph{BaseFold}-backed stack (E1 WHIR removal, April 2026).  The
BaseFold port matches SP1's \texttt{slop\_basefold} upstream;
jagged packing stays as the heterogeneous-batch adapter.  The new
per-chip variant (E3, in-progress) avoids the dense materialisation
the WHIR path required, eliminating a structural out-of-memory
failure mode and replacing a 2000\,LOC reduction with 16 focused
unit tests validating each primitive (LongMle, HadamardProduct,
sumcheck driver, BranchingProgram, and end-to-end single-chip +
multi-chip jagged-eval protocol runs).

All host-side code is implemented and tested in the open-source
Ziren codebase: the late-binding modules ($24$ WHIR-era tests
preserved as reference, passing), the recursion-circuit verifier
($18$ tests), zerocheck + LogUp-GKR ($9$ tests), and the BaseFold
stack ($56/56$ in \texttt{zkm-stark} including $16$ in
\texttt{jagged\_per\_chip}).  We document the workload-dependent
trade-off from WHIR profiling, motivate the migration, and report
BaseFold performance (fib-1k: $9.6$\,s prove, $362$\,ms verify,
$4.96$\,MB; keccak-precompile SDK: $50.9$\,s wall, $10.3$\,GB peak,
verified end-to-end).

\paragraph{Keywords:}
STARKs $\cdot$ multilinear polynomial commitment $\cdot$ WHIR
$\cdot$ BaseFold $\cdot$ jagged commitment $\cdot$ soundness $\cdot$
recursion $\cdot$ zero-knowledge virtual machines $\cdot$ Plonky3.
\end{abstract}

\tableofcontents
\vspace{1em}
\hrule
\vspace{1em}

% =============================================================================
% Section 1: Introduction
\input{sections/introduction}

% =============================================================================
% Section 2: Preliminaries
\input{sections/preliminaries}

% =============================================================================
% Section 3: Background and Motivation (status / context)
\input{sections/background}

% =============================================================================
% Section 4: Plonky3 Upgrade
\input{sections/plonky3-upgrade}

% =============================================================================
% Section 5: WHIR Fast Path Threat Model
\input{sections/soundness-fix}

% =============================================================================
% Section 6: System Design --- proof-system architecture (jagged + WHIR + late-
% binding + cross-binding composition, with implementation pointers)
\input{sections/design}

% =============================================================================
% Section 7: Phase 2a Construction --- Zerocheck for Transition Constraints
\input{sections/zerocheck-integration}

% =============================================================================
% Section 7: Phase 2b Construction --- LogUp-GKR with Sumcheck Reduction
\input{sections/logup-gkr-fix}

% =============================================================================
% Section 8: Phase 2c Construction --- Late-Binding WHIR + Jagged + Cross-Binding
\input{sections/whir-late-binding}

% =============================================================================
% Section 9: Performance Evaluation
\input{sections/performance}

% =============================================================================
% Section 10: Related Work
\input{sections/related-work}

% =============================================================================
% Section 11: Conclusion and Future Work
\input{sections/conclusion}

% =============================================================================
% Appendix: Implementation Status
\appendix
\input{sections/status-and-next-steps}
\input{sections/changelog}

% =============================================================================
% References
\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{whir2024}
U.\ Hab\"ock, D.\ Levit, S.\ Papini.
\emph{WHIR: Reed--Solomon Proximity Testing with Super-Fast Verification.}
IACR ePrint 2024/1586, 2024.

\bibitem{veil2026}
R.\ Dalal, T.\ Hemo, E.\ Rabinovich, R.~D.~Rothblum.
\emph{VEIL: Lightweight Zero-Knowledge for Hash-Based Multilinear Proof Systems.}
Succinct, March 2026.

\bibitem{logup-gkr}
A.\ Haböck.
\emph{Multivariate lookups based on logarithmic derivatives.}
IACR ePrint 2022/1530, 2022.

\bibitem{sumcheck}
C.\ Lund, L.\ Fortnow, H.\ Karloff, N.\ Nisan.
\emph{Algebraic methods for interactive proof systems.}
J.\ ACM 39(4), 1992.

\bibitem{brakedown}
A.\ Golovnev, J.\ Lee, S.\ Setty, J.\ Thaler, R.~S.~Wahby.
\emph{Brakedown: Linear-time and field-agnostic SNARKs for R1CS.}
CRYPTO 2023.

\bibitem{jagged}
T.\ Hemo et al.
\emph{Jagged Polynomial Commitments.}
IACR ePrint 2025/917, 2025.

\bibitem{plonky3}
The Plonky3 contributors.
\emph{Plonky3: A toolkit for STARK and SNARK proof systems.}
\url{https://github.com/Plonky3/Plonky3}, accessed 2026.

\bibitem{ziren}
ProjectZKM.
\emph{Ziren: A MIPS-based zero-knowledge virtual machine.}
\url{https://github.com/ProjectZKM/Ziren}, accessed 2026.

\bibitem{sp1-hypercube}
Succinct Labs.
\emph{SP1: A high-performance zkVM with hypercube architecture.}
\url{https://github.com/succinctlabs/sp1}, accessed 2026.

\end{thebibliography}

\end{document}
