\section{Background and Motivation}
\label{sec:background}

\subsection{The starting point}

Ziren's main branch ships a STARK prover with FRI as the polynomial
commitment scheme.  Each shard commits the main trace, the permutation
trace (for LogUp lookups), and a quotient polynomial (for transition
constraints), then opens all three at a univariate point $\zeta$.  The
verifier replays the same three opens and checks the AIR identities:
constraint-quotient relationship, lookup accumulation, cumulative-sum
closure.

This is the standard univariate-FRI STARK pipeline.  It has the
disadvantages typical of FRI: large per-shard proofs (megabytes for
RISC-V workloads), expensive recursion (many merkle paths per query),
and a large recursion circuit (the FRI verifier must be expressed in
the recursion AIR, with all the per-round PoW witnesses, log-arity
observations, and merkle-path checks).

\subsection{The Plonky3 / WHIR upgrade}

Two upstream developments motivated a major upgrade:

\begin{enumerate}
  \item \textbf{Plonky3 protocol updates.}  Upstream Plonky3 added
        per-round PoW witnesses, log-arity observations, and a
        restructured opened-values absorption pattern to the FRI
        verifier.  This is necessary for proper Fiat-Shamir binding
        with adaptive folding factors but breaks compatibility with
        any recursion circuit pinned to the older protocol.
  \item \textbf{WHIR multilinear PCS.}  The WHIR protocol
        \cite{whir2024} reduces multilinear polynomial commitments to
        a sequence of sumcheck rounds with folding, achieving
        $O(\log n)$ verifier work and proof sizes 5--10$\times$
        smaller than FRI for the same security level.  Plonky3's
        \texttt{p3-whir} adapter exposes WHIR as a
        \texttt{MultilinearPcs}.
\end{enumerate}

\subsection{The architecture of this work}

The work documented in this whitepaper does three things, in order:

\begin{enumerate}
  \item \textbf{Plonky3 upgrade.}  Bring Ziren's recursion circuit and
        FRI verifier in line with the new Plonky3 protocol.  Includes
        per-round PoW witnesses in the recursion FRI verifier, fixed
        \texttt{check\_witness} for empty PoW, and a \texttt{is\_div\_active}
        preprocessed column to gate \texttt{DivF}/\texttt{DivE} in the
        recursion AIR (see \S\ref{sec:plonky3-upgrade}).
  \item \textbf{Phase 2 soundness fix.}  Replace the quotient and
        permutation-trace mechanisms with sumcheck-based equivalents
        that work multilinearly: zerocheck (Phase 2a) and LogUp-GKR
        (Phase 2b).  This is what makes the WHIR fast path actually
        sound, not just a structural skip
        (see \S\ref{sec:zerocheck-phase2a}, \S\ref{sec:logup-gkr-fix}).
  \item \textbf{Late-binding WHIR adapter (Phase 2c).}  The remaining
        soundness gap after Phase 2b: per-chip row-MLE openings
        (\texttt{LogUpRowOpening}) are sent in the proof but not
        cryptographically tied to the main-trace commitment.  Closing
        this gap requires opening the WHIR commitment at multilinear
        points sampled \emph{after} the commit, which the upstream
        adapter's API forbids.  This work introduces a late-binding
        adapter that correctly handles post-commit point selection
        (see \S\ref{sec:whir-late-binding}).
\end{enumerate}

\subsection{Status of the work}

\begin{tabularx}{\textwidth}{@{}lXc@{}}
\toprule
\textbf{Item} & \textbf{Status} & \textbf{Tests} \\
\midrule
Plonky3 upgrade (recursion FRI verifier) & shipped, compressed proofs verify & --- \\
Cherry-pick aggregation fix (\texttt{7c2071e}) & shipped on \texttt{feat/upgrade-plonky3} & --- \\
Aggregation example with WHIR (ZK stages) & all stages succeed; gnark wrap blocked on stale artifacts & --- \\
Groth16 example with WHIR (ZK stages) & all stages succeed; gnark wrap blocked on stale artifacts & --- \\
Phase 2a (zerocheck) & shipped & 5/5 passing \\
Phase 2b (LogUp-GKR sumcheck reduction) & shipped & 9/9 passing \\
Phase 2c (single-column late-binding adapter) & shipped & 2/2 passing \\
Phase 2c (multi-column late-binding adapter) & shipped & 2/2 passing \\
Phase 2c (\texttt{LateBindingCapable} trait + KB impl) & shipped & --- \\
Phase 2c (full multi-chip wiring simulation) & shipped & 1/1 passing \\
Phase 2c (JSON serialisation + round-trip test) & shipped & 1/1 passing \\
Phase 2c (interleaved-codeword optimisation, VEIL Lemma 4.7) & designed; implementation deferred & --- \\
Prover wiring (\texttt{prover.rs} \texttt{try\_compute\_late\_binding\_proofs}) & shipped (TypeId dispatch, dual commit, dispatches both \texttt{KoalaBearPoseidon2} and \texttt{...Inner}) & --- \\
Verifier wiring (\texttt{verifier.rs} \texttt{try\_verify\_late\_binding\_proofs}) & shipped (TypeId dispatch, mirroring fresh challenger) & --- \\
End-to-end smoke test (fibonacci with \texttt{ZIREN\_USE\_WHIR=1}) & verified: 19 chips, 29.5\,MB late-binding bytes, 32.6\,s prover, verifier accepts & --- \\
Cross-binding (FRI $\leftrightarrow$ WHIR consistency) & not started; soundness gap documented & --- \\
Recursion circuit (\texttt{verify\_whir\_pcs}) & not started & --- \\
gnark BN254 wrap artifact regeneration & blocked on VK map regen & --- \\
\bottomrule
\end{tabularx}
