\section{LogUp-GKR: Bug, Fix, and End-to-End Integration}
\label{sec:logup-gkr-fix}

\subsection{The Soundness Bug (And Why It Was Hard to Spot)}

The initial Phase 2b implementation in
\texttt{crates/stark/src/logup\_gkr.rs} used a folding-only reduction.
At each GKR layer, the prover sent four values
\[
  (N_k(z \Vert 0),\; N_k(z \Vert 1),\; D_k(z \Vert 0),\; D_k(z \Vert 1)),
\]
and the verifier checked the combine identity
\begin{align*}
  N_{k+1}(z) &\stackrel{?}{=} N_k(z \Vert 0) \cdot D_k(z \Vert 1) + D_k(z \Vert 0) \cdot N_k(z \Vert 1), \\
  D_{k+1}(z) &\stackrel{?}{=} D_k(z \Vert 0) \cdot D_k(z \Vert 1),
\end{align*}
before folding a challenge $r^*$ to obtain the next claim at $z \Vert r^*$.

This identity is only valid for Boolean $z$. At a random $z \in \mathbb{F}^{m-k-1}$
the right-hand side is a degree-2 polynomial in each coordinate, while
$N_{k+1}$ is multilinear (degree 1). For arbitrary $z$ the identity fails.

\paragraph{Why the original test suite didn't catch it.}
The honest-roundtrip test used balanced fractions with all denominators
drawn from $\{7, 11\}$. Those repeated denominators made the degree-2
cross-terms vanish algebraically, so the check accidentally held. A
regression test with varied denominators
(\texttt{logup\_gkr\_roundtrip\_with\_varied\_denominators}) fails with
the folding-only reduction but now passes with the sumcheck-based fix.

\subsection{The Correct Reduction}

At each GKR layer we need to reduce a claim
$\bigl(N_{k+1}(z),\; D_{k+1}(z)\bigr)$ into claims on $N_k, D_k$ at a
fresh point. The MLE-extended combine identity
\[
  \widetilde{N}_{k+1}(z)
   \;=\; \sum_{b \in \{0,1\}^{m-k-1}} \mathsf{eq}(z, b) \cdot
         \bigl[\widetilde{N}_k(b, 0) \cdot \widetilde{D}_k(b, 1)
               + \widetilde{D}_k(b, 0) \cdot \widetilde{N}_k(b, 1)\bigr]
\]
is genuinely identical at every $z$ because the product of multilinear
functions is only evaluated at \emph{Boolean} $b$ inside the sum. We
turn this into a sumcheck claim.

\subsubsection*{Protocol per layer}
\begin{enumerate}
  \item \textbf{Batch.} Sample $\lambda \in \mathbb{F}_{\mathrm{ext}}$. The
        merged claim is $\lambda \cdot \widetilde{N}_{k+1}(z) + \widetilde{D}_{k+1}(z)$.
  \item \textbf{Sumcheck.} Run $m-k-1$ rounds of sumcheck over
        \[
          g(b) \;=\; \mathsf{eq}(z, b) \cdot
               \bigl(\lambda \cdot (\widetilde{N}_k \widetilde{D}_k \text{ cross})
                     + \widetilde{D}_k(b, 0) \cdot \widetilde{D}_k(b, 1)\bigr).
        \]
        $g$ has total degree 3 in each variable being summed out, so each
        round sends 4 field elements (evaluations at $X \in \{0, 1, 2, 3\}$).
  \item \textbf{Final evaluation.} The prover sends
        $(\widetilde{N}_k(r^*, 0),\; \widetilde{N}_k(r^*, 1),\;
          \widetilde{D}_k(r^*, 0),\; \widetilde{D}_k(r^*, 1))$. The
        verifier checks the sumcheck's final identity directly.
  \item \textbf{Line protocol.} Sample $t \in \mathbb{F}_{\mathrm{ext}}$
        and fold the two pairs of values into a single claim at $(r^*, t)$.
\end{enumerate}

After $m$ layers, the accumulated point $\mathbf{r} \in \mathbb{F}^m$ and
the final claim $(\widetilde{N}_0(\mathbf{r}), \widetilde{D}_0(\mathbf{r}))$
can be checked against the fingerprints reconstructed from the
main-trace PCS opening at $\mathbf{r}$.

\subsection{Implementation}

\paragraph{Proof shape.} The new \texttt{LogUpGkrProof} carries per-layer
reductions:
\begin{verbatim}
pub struct LogUpGkrLayerProof<EF> {
    pub sumcheck_rounds: Vec<[EF; 4]>, // degree-3 round polys
    pub final_evals: [EF; 4],          // (N_k(r*,0), N_k(r*,1), D_k(r*,0), D_k(r*,1))
}

pub struct LogUpGkrProof<EF> {
    pub root: (EF, EF),
    pub layers: Vec<LogUpGkrLayerProof<EF>>,
    pub eval_point: Vec<EF>,
    pub leaf_claim: (EF, EF),
}
\end{verbatim}

\paragraph{Variable ordering.} The sumcheck folds the last variable
first at every round (cache-friendly pair-consecutive layout), so the
raw \texttt{r\_star} vector built during the sumcheck encodes challenges
in reverse variable order. The prover and verifier both reverse
\texttt{r\_star} before appending the line challenge, so the returned
\texttt{eval\_point} matches the \texttt{eval\_mle\_first\_var}
convention used by the caller.

\paragraph{Unit tests.} All 8 tests pass:
\begin{itemize}
  \item \texttt{fraction\_combine\_is\_associative\_on\_small\_input}
  \item \texttt{sum\_of\_balanced\_fractions\_has\_zero\_numerator}
  \item \texttt{logup\_gkr\_prove\_verify\_roundtrip\_on\_balanced\_input}
  \item \texttt{logup\_gkr\_leaf\_claim\_matches\_mle\_evaluation}
        \textit{(final claim equals MLE at \texttt{eval\_point})}
  \item \texttt{logup\_gkr\_roundtrip\_with\_varied\_denominators}
        \textit{(regression for the folding-only bug)}
  \item \texttt{logup\_gkr\_rejects\_tampered\_final\_evals}
  \item \texttt{logup\_gkr\_rejects\_tampered\_root}
  \item \texttt{logup\_gkr\_leaf\_claim\_reconstructs\_from\_row\_mle}
        \textit{(end-to-end: real chip $\to$ leaves $\to$ GKR $\to$
        reconstruction from row-MLE opening matches)}
\end{itemize}

\paragraph{End-to-end.} The WHIR fast path now emits LogUp-GKR proofs
per chip and the verifier replays them. Fibonacci core proof: 7.2\,s
(1.15\,kHz); compressed: 7.1\,s.

\begin{table}[h]
\centering
\begin{tabular}{lrrl}
\toprule
\textbf{Mode} & \textbf{e2e (s)} & \textbf{kHz} & \textbf{Status} \\
\midrule
main FRI            & 12.54 & 0.66 & Baseline \\
upgrade FRI         & 8.21  & 1.01 & +53\% vs main \\
upgrade WHIR (zerocheck skipped, no GKR) & 6.8  & 1.22 & +85\% vs main, unsound \\
upgrade WHIR + sumcheck LogUp-GKR        & 7.2  & 1.15 & +74\% vs main, transcript-bound GKR \\
\bottomrule
\end{tabular}
\caption{Fibonacci core-proof times across configurations. LogUp-GKR
adds ~5\% over the bare WHIR path thanks to the 1.5\,s per-shard sumcheck
cost. Compared to main FRI, we keep 74\% of the measured speedup while
restoring lookup-argument binding.}
\label{tab:logup-gkr-perf}
\end{table}

\subsection{Closing the Soundness Loop}

The \texttt{reconstruct\_leaf\_claim\_from\_openings} helper
(\texttt{crates/stark/src/logup\_gkr.rs}) implements the final soundness
check.  Given the opened main-trace and preprocessed-trace values at the
row coordinates of $\mathbf{r}$, it evaluates each lookup interaction's
multiplicity and fingerprint expressions at that row point (using
\texttt{VirtualPairCol::apply::<EF, EF>}), then combines the
per-interaction fractions into the leaf-layer MLE values at
$\mathbf{r}$:
\begin{align*}
  \widetilde{N}_0(\mathbf{r})
    &= \sum_{j} \mathsf{eq}(\mathbf{r}_{\mathrm{int}}, j) \cdot \mathrm{mult}_j(\mathbf{r}_{\mathrm{row}}), \\
  \widetilde{D}_0(\mathbf{r})
    &= \sum_{j} \mathsf{eq}(\mathbf{r}_{\mathrm{int}}, j) \cdot \bigl(\alpha + \beta \cdot \mathrm{kind}_j + \textstyle\sum_i \beta^{i+1} \cdot \mathrm{value}_{j,i}(\mathbf{r}_{\mathrm{row}})\bigr).
\end{align*}

The unit test \texttt{logup\_gkr\_leaf\_claim\_reconstructs\_from\_row\_mle}
builds a real single-column chip with a single lookup, runs the GKR
prover, evaluates the row-MLE of the trace at \texttt{proof.eval\_point}
using \texttt{eval\_mle\_first\_var}, calls the reconstruction helper,
and confirms the result matches \texttt{proof.leaf\_claim} exactly ---
closing the protocol's soundness chain end-to-end on synthetic data.

\paragraph{Outstanding infrastructure.} What the verifier still lacks is
the ability to \emph{obtain} the main-trace row-MLE values at
\texttt{proof.eval\_point}: today the WHIR fast path opens the trace
only at a univariate point $\zeta$.  Once the PCS layer exposes
multilinear opening at arbitrary points (the same infrastructure needed
to close the Phase~2a zerocheck gap), the verifier will:
\begin{enumerate}
  \item After \texttt{verify\_logup\_gkr}, obtain the opened trace
        values at \texttt{proof.eval\_point} via WHIR.
  \item Split \texttt{eval\_point} into \texttt{(r\_row, r\_int)}.
  \item Call \texttt{reconstruct\_leaf\_claim\_from\_openings}.
  \item Compare the reconstruction to \texttt{proof.leaf\_claim}.
\end{enumerate}
All of the arithmetic is already implemented and unit-tested; only the
PCS-level plumbing remains.

\subsection{Files Touched}

\begin{verbatim}
crates/stark/src/logup_gkr.rs         rewritten with sumcheck reduction
crates/stark/src/zerocheck_prover.rs  made fold_table_first pub
crates/stark/src/prover.rs            WHIR path emits LogUpGkrProof per chip
crates/stark/src/verifier.rs          WHIR path replays LogUpGkrProof
crates/stark/src/types.rs             ShardProof::logup_gkr_proofs field
crates/sdk/src/provers/mock.rs        literal update
crates/recursion/circuit/src/stark.rs literal update
\end{verbatim}
