\section{Plonky3 Upgrade}
\label{sec:plonky3-upgrade}

The recursion circuit needed several fixes to work with the new
upstream FRI verifier.  All fixes are committed on
\texttt{feat/upgrade-plonky3}.

\subsection{Per-round PoW witnesses}

Upstream's new FRI requires the prover to grind a proof-of-work
witness once per round (not just at the final query stage).  The
recursion-circuit FRI verifier
(\texttt{crates/recursion/circuit/src/fri.rs}) was missing the
per-round \texttt{check\_witness} call.  Added.

\subsection{Log-arity observations and opened-values restoration}

The new FRI verifier observes the log-arity of each round and absorbs
the opened-values vector into the transcript before sampling the next
fold randomness.  Both observations were stripped in an earlier
"compatibility" patch and were restored in
\texttt{p3-whir} commit \texttt{715fdb5}.  The recursion circuit
mirrors them.

\subsection{\texttt{check\_witness} early-return for zero PoW bits}

\texttt{GrindingChallenger::check\_witness} returns \texttt{true} when
the requested bits is zero.  The recursion circuit's variable-arity
version was running the hash anyway; aligned to native behaviour.

\subsection{\texttt{is\_div\_active} preprocessed column}

The recursion compiler emits \texttt{DivF}/\texttt{DivE} instructions
in both branches of \texttt{Select}, including dead branches where the
divisor is zero.  Native execution handles this gracefully because
\texttt{mult==0} skips the operation.  The AIR didn't.  Added a
preprocessed column \texttt{is\_div\_active = is\_div AND
mult\_active} (where \texttt{mult\_active} marks active alu rows) and
gated the division constraint on it.  Mirror runtime guard added in
\texttt{crates/recursion/core/src/runtime/mod.rs}.

\subsection{Result}

Compressed proofs for fibonacci and other benchmarks verify with the
upgraded recursion circuit.  Aggregation runs through
\texttt{prove\_core $\to$ compress $\to$ shrink $\to$ wrap\_bn254}
successfully; the gnark BN254 wrap fails with witness-count mismatch
because the gnark circuit artifacts predate the FRI changes (250
extra witness elements; see \S\ref{sec:performance}).
