\section{Soundness Fix Design: Making WHIR Mode Production-Sound}
\label{sec:soundness-fix}

\subsection{Current State}

The WHIR fast path (\texttt{ZIREN\_USE\_WHIR=1}) in
\texttt{crates/stark/src/prover.rs} skips transition-constraint verification
entirely. The core prover:
\begin{enumerate}
  \item Commits to the main trace only (no permutation trace, no quotient).
  \item Opens the main trace at $\zeta$ via WHIR (Jagged).
  \item Sets \texttt{ShardCommitment.permutation\_commit = None} and
        \texttt{quotient\_commit = None}.
  \item Leaves \texttt{ChipOpenedValues.permutation} and \texttt{quotient}
        empty.
\end{enumerate}

The recursion verifier (both native and circuit) short-circuits when
\texttt{permutation\_commit.is\_none()}: it skips the PCS round for
permutation/quotient, skips \texttt{verify\_constraints}, and only checks:
\begin{itemize}
  \item PCS opening of preprocessed + main.
  \item $\sum_i \text{local\_cumulative\_sum}_i = 0$.
\end{itemize}

\subsection{What This Leaves Unchecked}

Given a committed main trace $M$ and cumulative-sum extraction from the last
14 base-field columns of $M$, the verifier does \emph{not} check:

\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{AIR check} & \textbf{FRI mode} & \textbf{WHIR fast path} \\
\midrule
$\sum \text{lookup events} = 0$ (permutation argument) &
  \checkmark\ via quotient check &
  \checkmark\ via cumulative sum \\
Transition: $C(M[i], M[i+1], P[i]) = 0 \ \forall i$ &
  \checkmark\ via quotient check &
  $\times$\ NOT CHECKED \\
Boundary: $C_{\text{first}}(M[0]) = 0$, $C_{\text{last}}(M[n{-}1]) = 0$ &
  \checkmark\ via Lagrange selectors &
  $\times$\ NOT CHECKED \\
Public values = $M[0][\text{pub\_cols}]$ &
  \checkmark\ via boundary constraint &
  $\times$\ NOT CHECKED \\
\bottomrule
\end{tabular}
\caption{What is verified in each mode.}
\label{tab:soundness-gap}
\end{table}

A malicious prover can therefore produce a ``verifying'' proof for a main
trace that does not come from any valid execution of the MIPS program,
forge public outputs, or violate transition constraints. Only the global
septic digest (extracted from the last 14 columns) is bound, and that is
bindable separately to memory-init/finalize events.

\emph{The current WHIR mode is a benchmark prototype, not a zk-proof.}

\subsection{Target Soundness Design}

The fix must re-introduce transition-constraint verification while
preserving the Jagged + WHIR performance characteristics. There are three
standard options.

\paragraph{Option A --- Keep the quotient polynomial, batch it into WHIR.}
Commit to the quotient polynomial alongside the main trace, open both at
$\zeta$, check $q(\zeta) \cdot Z_H(\zeta) = \text{folded\_constraints}(\zeta)$.
\emph{Pro:} minimal change. \emph{Con:} defeats most of the WHIR speedup ---
quotient width is $2^{\text{log\_quotient\_degree}}$ (= 2 in Ziren), each
column must be computed, committed, and opened. The tendermint $4.5\times$
speedup would collapse to $\sim$1.3$\times$. \emph{Status:} not recommended.

\paragraph{Option B --- Logup-GKR plus sumcheck-based constraint verification
(SP1 hypercube model).}
Replace both the permutation argument and the quotient polynomial with GKR /
sumcheck reductions:
\begin{enumerate}
  \item \textbf{Constraints $\to$ sumcheck.} The transition-constraint
        polynomial $C(X) = \alpha \cdot C_0(X) + \alpha^2 \cdot C_1(X) + \cdots$
        is viewed as a multilinear polynomial over the Boolean hypercube
        $\{0,1\}^{\log n}$. The prover proves
        $$
          \sum_{x \in \{0,1\}^{\log n}} \mathsf{eq}(r, x) \cdot C(x, x_{\text{next}}) = 0
        $$
        via sumcheck ($\log n$ rounds, $\sim$3 field elements per round).
  \item \textbf{Lookups $\to$ Logup-GKR.} The lookup identity becomes a GKR
        circuit whose final claim is a multilinear evaluation of a fraction
        polynomial. The prover runs a few sumcheck rounds to reduce it to a
        single evaluation.
  \item \textbf{Final opening.} All final claims are batched and opened at
        one multilinear point via WHIR.
\end{enumerate}
\emph{Pro:} matches SP1's direction; Logup-GKR has battle-tested designs;
preserves most of the WHIR speedup because no extra trace is committed.
\emph{Con:} large implementation effort (requires sumcheck-verifier chips
inside the recursion circuit). \emph{Status:} recommended (this document's
focus).

\paragraph{Option C --- Multilinear constraint check via WHIR opening at $\zeta$.}
Keep the univariate STARK constraint check but express it in multilinear
form. The verifier:
\begin{enumerate}
  \item Samples $\alpha, \zeta$.
  \item Opens $\text{main}(\zeta), \text{main}(\zeta \omega), \text{preproc}(\zeta), \text{preproc}(\zeta \omega)$ via WHIR.
  \item Locally computes $\text{folded\_constraints}(\zeta)$ from the opened values.
  \item Asserts $\text{folded\_constraints}(\zeta) = 0$.
\end{enumerate}
The last step replaces the identity $q(\zeta) \cdot Z_H(\zeta) = \text{folded}(\zeta)$
with the direct assertion $\text{folded}(\zeta) = 0$. This is only sound if
$\text{folded\_constraints}(X)$ vanishes on the trace domain, which still
requires a sumcheck argument to prove --- collapsing Option C into
Option B. \emph{Status:} subsumed by Option B.

\subsection{Chosen Design: Option B}

\subsubsection*{Protocol sketch (prover)}

Let $M$ be the main trace (multilinear, dimension $\log n$ over
$\{0,1\}^{\log n}$), $P$ the preprocessed trace, $\mathrm{pub}$ the public
values.

\begin{enumerate}[label=\textbf{Round \arabic*.}, wide=0pt]
  \item Commit to $M$ via WHIR (same as today).
  \item Sample constraint-mixing challenge $\alpha$.
  \item \textbf{Constraint sumcheck.} Prover and verifier run sumcheck for
        $$
          \sum_{x \in \{0,1\}^{\log n}} \mathsf{eq}(r, x) \cdot \Bigl[ \sum_i \alpha^i \cdot C_i\bigl(M(x), M(x \omega), P(x), P(x \omega), \mathrm{pub}\bigr) \Bigr] = 0,
        $$
        where $r \in \mathbb{F}^{\log n}$ is sampled by the verifier at the
        start. After $\log n$ rounds, the claim reduces to a single
        multilinear-value claim on $M$ and $P$ at point $r' \in \mathbb{F}^{\log n}$.
  \item \textbf{Lookup (Logup-GKR).} For each lookup table $T$ with events
        $e_1, \dots, e_k$:
        $$
          \sum_j \frac{1}{\beta - e_j} = \sum_t \frac{\text{multiplicity}(t)}{\beta - t}.
        $$
        The prover runs a GKR circuit whose output wire is $0$. The GKR
        reduction ends with a claim on $M, P$ at another point $r''$.
  \item \textbf{Batched WHIR opening.} All multilinear claims on $M$ (from
        sumchecks and GKR) are batched via random linear combination and
        opened at a single point via WHIR. Same for $P$.
\end{enumerate}

\subsubsection*{Verifier cost}

\begin{itemize}
  \item Sumcheck of degree-$d$ polynomial over $\log n$ rounds:
        $\log n \cdot (d+1)$ field elements observed, one final multilinear
        evaluation check.
  \item Logup-GKR: $O(\log k)$ sumcheck rounds per table.
  \item WHIR opening: same cost as today's Jagged + WHIR open.
\end{itemize}
Total verifier work: $O\bigl(\log n \cdot (d + |\text{tables}|)\bigr)$ ---
roughly the same asymptotic cost as today, \emph{plus} constraint
soundness.

\subsubsection*{Recursion implications}

The recursion circuit currently contains:
\begin{itemize}
  \item \texttt{verify\_shard} --- iterates over chips, calls
        \texttt{verify\_constraints}.
  \item \texttt{verify\_constraints} --- computes
        $\text{folded\_constraints}(\zeta)$ and
        $q(\zeta) \cdot Z_H(\zeta)$ and asserts equality.
\end{itemize}

Under Option B the recursion circuit needs:
\begin{enumerate}
  \item A \texttt{SumcheckVerify} chip (skeleton already exists in
        \texttt{crates/recursion/core/src/chips/sumcheck\_verify.rs}).
  \item A \texttt{GKRVerifier} circuit (new).
  \item A \texttt{MultilinearEvalCheck} that reconciles a claimed
        multilinear evaluation at $r$ with the value opened by WHIR.
  \item A replacement for \texttt{verify\_constraints} that runs
        \texttt{SumcheckVerify} per chip, runs \texttt{GKRVerifier} per
        lookup table, batches all multilinear claims, and checks them
        against WHIR-opened values.
\end{enumerate}

\subsubsection*{Migration path}

\begin{enumerate}
  \item \textbf{Phase 2a (done)} --- sumcheck primitives, hypercube
        constraint evaluator, \texttt{ShardProof::zerocheck\_proofs} field,
        prover + verifier wiring. See Section~\ref{sec:zerocheck-phase2a}.
  \item \textbf{Phase 2b} --- Logup-GKR for lookup arguments in the native
        prover. Removes the permutation trace entirely.
  \item \textbf{Phase 2c} --- wire \texttt{SumcheckVerify} and
        \texttt{GKRVerifier} into the recursion circuit. Replace
        \texttt{verify\_constraints} with the sumcheck-based version.
  \item \textbf{Phase 3} --- formal verification of the sumcheck / GKR
        verifier chips.
\end{enumerate}

\subsection{Projected Performance Impact}

Compared to the current (unsound) WHIR fast path:

\begin{table}[h]
\centering
\small
\begin{tabular}{lllc}
\toprule
\textbf{Operation} & \textbf{WHIR fast path} & \textbf{WHIR + Logup-GKR} & \textbf{Overhead} \\
\midrule
Main trace commit   & same   & same    & 0\%       \\
Constraint check    & skipped & sumcheck ($\log n$ rounds) & +10--15\% \\
Lookup check        & cumulative sum read & Logup-GKR sumcheck & +5--10\%  \\
PCS open            & one point & batched multi-point & +5\%      \\
\midrule
\textbf{Total expected} & \textbf{baseline} & \multicolumn{2}{r}{\textbf{+20--30\%}} \\
\bottomrule
\end{tabular}
\caption{Projected end-to-end overhead of Option B relative to the unsound
WHIR fast path.}
\label{tab:sound-overhead}
\end{table}

Projected tendermint performance: $502\,\text{s} \cdot 1.25 \approx 630\,\text{s}$,
still $\mathbf{3.6\times}$ faster than main FRI (2270\,s), with full
soundness restored.

\subsection{Immediate Action Items}
\begin{enumerate}
  \item Rename the current flag to \texttt{ZIREN\_WHIR\_UNSAFE=1} with a
        runtime warning; the existing name misleads about soundness.
  \item Complete Phase 2a follow-up: multi-point WHIR opening so the
        sumcheck final claim $C(r')$ can be checked against an actual
        opening.
  \item Implement Phase 2b (Logup-GKR) to make all chips zerocheck-eligible.
  \item Wire into the recursion circuit (Phase 2c).
  \item Formal-verification pass (Phase 3).
\end{enumerate}
