\section{Implementation Status and Next Steps}
\label{sec:status-next}

\subsection{What ships today}

\begin{itemize}
  \item Plonky3 upgrade complete; compressed proofs verify.
  \item Phase 2a zerocheck integration: 5/5 tests passing.
  \item Phase 2b LogUp-GKR with proper sumcheck reduction: 9/9 tests
        passing, including end-to-end leaf-claim reconstruction.
  \item Phase 2c late-binding WHIR adapter (single + multi-column,
        per-column WHIR commits): 4/4 tests passing including
        tampering rejection.
  \item Cherry-pick of aggregation fix \texttt{7c2071e} from \texttt{main}.
  \item Aggregation and Groth16 examples run in WHIR mode through
        \texttt{wrap\_bn254}; final gnark wrap blocked on stale
        artifact regeneration (independent of soundness work).
\end{itemize}

\subsection{What's next}

\begin{enumerate}
  \item \textbf{Wire the late-binding adapter into
        \texttt{crates/stark/src/prover.rs}.}
        Replace per-chip \texttt{pcs.commit} / \texttt{pcs.open} with
        \texttt{commit\_multi\_column} / \texttt{add\_late\_claim\_multi\_column}
        / \texttt{open\_multi\_column}.  After GKR derives
        $r_{\mathrm{row}}$, register the per-column row-MLE values
        from \texttt{LogUpRowOpening.main\_at\_r\_row} as late claims
        and run the open protocol.
  \item \textbf{Wire verifier symmetric in
        \texttt{crates/stark/src/verifier.rs}.}
        Parse the new opening proof per chip
        (\texttt{Vec<WhirProof>}), replay the late binding with
        \texttt{replay\_multi\_column\_commits} and
        \texttt{finish\_verify\_multi\_column}, and reuse the existing
        \texttt{reconstruct\_leaf\_claim\_from\_openings} unchanged
        because the per-column row-MLE values it consumes are now
        cryptographically bound.
  \item \textbf{Optimise to interleaved-codeword construction.}
        Replace per-column commits with VEIL Lemma 4.7's interleaved-
        codeword + post-commit-$\rho$ design (see
        \S\ref{sec:whir-late-binding}).  Reduces per-chip proof size
        from $w$ WHIR proofs to one base WHIR proof plus $w$ exposed
        field elements.  Caller-visible API is unchanged.
  \item \textbf{Recursion circuit.}  Update
        \texttt{verify\_whir\_pcs} in
        \texttt{crates/recursion/circuit/src/whir\_verifier.rs} to
        consume the new opening proof shape.  Required for end-to-end
        compressed-proof verification with the late-binding wrapper.
  \item \textbf{Regenerate gnark BN254 wrap artifacts.}  Independent
        of soundness work; required for the final
        \texttt{Plonk}/\texttt{Groth16} wrap stages to succeed.
        Currently blocked on stale VK map regeneration
        (\texttt{vk not allowed} error in
        \texttt{build\_plonk\_bn254}).
\end{enumerate}

\subsection{Soundness summary}

After Phase 2c full wiring lands:

\begin{itemize}
  \item Transition constraints are bound by zerocheck.
  \item Lookup arguments are bound by LogUp-GKR sumcheck.
  \item Per-chip row-MLE openings are bound by per-column WHIR
        late-binding (or its interleaved-codeword optimisation).
  \item No remaining structural skips in the WHIR fast path.
\end{itemize}

WHIR mode then matches FRI mode in soundness, with smaller proofs and
multilinear-native PCS structure.
