\section{Phase 2c: WHIR Late-Binding and Per-Chip Row-MLE}
\label{sec:whir-late-binding}

Phase 2c closes the soundness gap left by Phases 2a (zerocheck) and 2b
(LogUp-GKR): the per-chip \texttt{LogUpRowOpening} values are bound
cryptographically to the main-trace WHIR commitment.  This requires
two pieces of new infrastructure: a WHIR PCS adapter that supports
\emph{late-binding} evaluation points, and a restructured GKR leaf
claim that needs only a single $\alpha$-combined row-MLE value per
chip rather than one per column.

\subsection{The Problem with the Standard WHIR Adapter}

The upstream \texttt{MultilinearPcs::commit(poly, opening\_points, ch)}
adapter requires the evaluation points at commit time: opening points
are absorbed into the Fiat-Shamir transcript \emph{before} the
column-batching challenge $\alpha$ is sampled.  This binding order is
essential for the upstream soundness proof.

Our protocol cannot satisfy that ordering.  After committing the main
trace we run zerocheck (which derives $\zeta$) and LogUp-GKR (which
derives a per-chip evaluation point $r_{\mathrm{row}}$); only then do
we know the full set of points at which we want to open the trace.
Sampling $r_{\mathrm{row}}$ before commit is impossible because GKR's
sumcheck transcript depends on the merkle root.

\subsection{Late-Binding Adapter}

\paragraph{Insight.}
WHIR's commitment phase already adds OOD constraints to the equality
statement \emph{after} the merkle root is observed
(\texttt{committer.commit} in \texttt{p3-whir}).  The cryptography
permits adding constraints between merkle commit and the WHIR rounds;
only the upstream API hides this capability behind a one-shot
\texttt{commit} call.

\paragraph{Implementation.}
\texttt{crates/stark/src/whir\_late\_binding.rs} (single column):
\begin{itemize}
  \item \texttt{commit\_column(poly, ch) -> (commit, prover\_state)}\\
        domain separator $\to$ width $\to$ merkle commit, observe root
        $\to$ OOD sample/observe.
  \item \texttt{add\_late\_claim(state, point, ch) -> value}\\
        evaluate $p(\mathrm{point})$, add to statement, observe
        \emph{value} into challenger.  Point itself is not re-observed:
        it is already bound by the caller's downstream protocol that
        derived it.
  \item \texttt{open(state, ch) -> (proof, late\_claims)}\\
        runs WHIR's prove phase against the accumulated statement.
  \item \texttt{replay\_commit\_into\_challenger(proof, ch) -> parsed}\\
        verifier-side: re-runs the commit-phase challenger
        interactions to reach the same state the prover had.
  \item \texttt{finish\_verify(proof, parsed, late\_claims, ch) -> Result}\\
        verifier-side: builds the equality statement in the same order
        the prover did (OOD first, then late claims), passes an
        empty-OOD \texttt{ParsedCommitment} to suppress the internal
        \texttt{concatenate} (which would re-add OOD constraints in
        the wrong position and corrupt $\gamma$-batching).
\end{itemize}

\paragraph{Soundness validation.}
Two unit tests in \texttt{whir\_late\_binding.rs}:
\begin{itemize}
  \item \texttt{late\_binding\_round\_trip\_single\_column}: end-to-end
        commit $\to$ derive point $\to$ add claim $\to$ prove $\to$
        verify with prover and verifier sampling the late point from
        their own challengers (sanity-checked via \texttt{assert\_eq!}).
  \item \texttt{late\_binding\_rejects\_tampered\_value}: confirms the
        binding is cryptographic --- a tampered late-claim value is
        rejected by \texttt{finish\_verify} (returns
        \texttt{Sumcheck(InvalidPowWitness)}).
\end{itemize}

\paragraph{Constraint ordering subtlety.}
\texttt{combine\_hypercube} weights the $i$-th constraint by
$\gamma^i$.  The prover's statement is
$[\mathrm{ood}_0, \ldots, \mathrm{ood}_{N-1}, \mathrm{late}_0,
\ldots, \mathrm{late}_{L-1}]$, set by
\texttt{CommitmentWriter.commit} (OOD first) then
\texttt{add\_late\_claim} (late after).  The naive verifier would call
\texttt{Verifier::verify(parsed, public\_statement = late\_claims)},
which internally does \texttt{statement.concatenate(\&parsed.ood)} and
ends up with $[\mathrm{late}, \mathrm{ood}]$ --- different
$\gamma^i$-weights, transcript mismatch, sumcheck rejects.  Our
\texttt{finish\_verify} builds the full $[\mathrm{ood}, \mathrm{late}]$
statement explicitly and passes an empty-OOD \texttt{ParsedCommitment}
to neutralise the internal concatenation.

\subsection{Architectural Choice for Multi-Column Binding}

Each chip's main trace has $w$ columns.  We need to bind the
per-column row-MLE evaluations $v_j = \mathsf{MLE}(\mathrm{col}_j)
(r_{\mathrm{row}})$ in the chip's \texttt{LogUpRowOpening} to the
WHIR commitment of the chip's trace.

Two viable approaches:

\paragraph{(A) Per-column WHIR commits.}
Commit each of the $w$ columns separately, run the late-binding
adapter $w$ times.  Soundness clean: each $v_j$ is bound by its own
WHIR opening at $r_{\mathrm{row}}$.  Cost: $w \times \#\mathrm{chips}$
merkle trees and WHIR proofs.  For a typical RISC-V STARK
($w \approx 30$, $\#\mathrm{chips} \approx 100$) this is $\sim 3000$
WHIR commits per shard --- a $30\times$ overhead vs the existing
single-batch commit.

\paragraph{(B) Single $\alpha$-combined commit + restructured GKR.}
Commit each chip's trace as a single multi-column matrix.  WHIR
internally batches columns via $g_c = \sum_{j=0}^{w-1} \alpha_c^j
\cdot \mathrm{col}_{c,j}$ where $\alpha_c$ is sampled at commit time
for chip $c$.  Restructure LogUp-GKR so its leaf-claim reconstruction
needs only the \emph{combined} row-MLE
$V_c := g_c(r_{\mathrm{row}}) = \sum_j \alpha_c^j \cdot v_{c,j}$
rather than the individual $v_{c,j}$.  Cost: one WHIR commit per
chip, matching the existing structure.

\paragraph{Why (B) is sound.}
The malicious-prover attack against the naive multi-column late
binding is: prover sees $\alpha_c$ before $r_{\mathrm{row}}$, then
chooses per-column values $v_{c,j}$ to satisfy
$\sum_j \alpha_c^j \cdot v_{c,j} = g_c(r_{\mathrm{row}})$, gaining
$w-1$ degrees of freedom.  Approach (B) eliminates this attack by
\emph{never trusting the prover with per-column values}: only the
combined value $V_c$ is needed for verification, and $V_c$ is bound
directly by WHIR's opening proof.  No $w-1$ slack exists because
there is no $w$-dimensional unknown.

\paragraph{Why we choose (B).}
Soundness is identical to (A); cost is $30\times$ smaller; the
existing chip commit shape carries through with no per-column
explosion.  The price is a non-trivial change to the LogUp-GKR
leaf-claim reconstruction.

\subsection{Restructured Leaf-Claim Reconstruction}

Current (Phase 2b) leaf-claim shape (\texttt{logup\_gkr.rs}):
$$
(N_{\mathrm{leaf}}, D_{\mathrm{leaf}}) =
\bigoplus_i \bigl(m_i,\ \alpha - f_i(\mathrm{row})\bigr)
$$
where $m_i$ is the multiplicity and $f_i$ is the fingerprint of the
$i$-th interaction at row $\mathrm{row}$.  Each $f_i$ is a polynomial
in the per-column values $v_{c,j} = \mathsf{MLE}(\mathrm{col}_{c,j})
(r_{\mathrm{row}})$ of the chip's main and preprocessed columns
(typically $f_i(\mathrm{row}) = \beta + \mathrm{kind}_i + \sum_k
\beta^k \cdot v_{c,j_{i,k}}$).

The reconstruction
\texttt{reconstruct\_leaf\_claim\_from\_openings} currently:
\begin{enumerate}
  \item Reads $v_{c,j}$ from \texttt{LogUpRowOpening.main\_at\_r\_row}
        and \texttt{preproc\_at\_r\_row}.
  \item For each interaction $i$, computes $f_i$ from the relevant
        $v_{c,j}$.
  \item Folds the $L$ interactions through the fraction-sum tree to
        get $(N_{\mathrm{leaf}}, D_{\mathrm{leaf}})$.
\end{enumerate}

Restructured (Phase 2c) leaf-claim shape:
$$
V_c := \sum_{j=0}^{w_c-1} \alpha_c^j \cdot v_{c,j}
\quad\text{(single per-chip combined value, bound by WHIR)}
$$
$$
V_c^{\mathrm{prep}} := \sum_{j=0}^{w'_c-1} \alpha_c^{w_c+j} \cdot v_{c,j}^{\mathrm{prep}}
\quad\text{(combined preprocessed, same chip's $\alpha_c$)}
$$

The reconstruction now needs to compute $f_i$ given only $V_c$ and
$V_c^{\mathrm{prep}}$ (and the public per-interaction shape).  This
is impossible in general: $f_i$ depends on \emph{specific} columns
$v_{c,j_{i,k}}$, not their $\alpha_c$-mix.

\subsubsection{The Fix: Fingerprint Re-Batching at GKR Time}

Restructure the GKR leaves so each leaf $\ell$ is itself a single
$\beta$-combined value carrying enough information that the
verifier can reconstruct it from $V_c$ alone.

Concretely, fold the per-interaction fingerprint into the column
batching: instead of computing $f_i$ from individual $v_{c,j}$, we
choose the column-batching coefficients such that
$$
f_i = \beta_i \cdot V_c + \mathrm{(public terms)}
$$
where $\beta_i$ is fixed by the protocol's interaction shape and is
the same on both sides.  The challenge is that different interactions
$i$ touch different subsets of columns; mapping all of them into a
single $V_c$ requires redefining the column-batching coefficients
per-interaction.

\paragraph{Concrete construction.}
For chip $c$ with $L_c$ interactions and total column count
$w_c + w'_c$ (main + preprocessed), define:
\begin{itemize}
  \item Per-interaction column-projection vectors
        $\pi_i \in \mathbb{F}^{w_c + w'_c}$ where $\pi_i[j] = \beta^k$
        if interaction $i$ uses column $j$ in slot $k$, else $0$.
  \item A single chip-wide combined polynomial
        $g_c(\mathbf{x}) = \sum_i \gamma^i \cdot \langle \pi_i,
        \mathrm{cols}(\mathbf{x}) \rangle$
        where $\gamma$ is sampled from the post-merkle challenger and
        $\mathrm{cols}(\mathbf{x})$ is the column vector of the chip
        evaluated at $\mathbf{x}$.
  \item Bind $g_c(r_{\mathrm{row}})$ via WHIR late-binding.
\end{itemize}

The verifier reconstructs the leaf claim from a \emph{single}
$g_c(r_{\mathrm{row}})$ value.  The per-interaction fingerprint
contributions are stitched together via $\gamma^i$ binding, which is
sampled \emph{after} $r_{\mathrm{row}}$ is known --- so all degrees
of freedom in per-column values are eliminated by the
$\gamma$-randomness.

\paragraph{Soundness sketch.}
A malicious prover wishing to lie about a leaf-claim value would need
to produce a $g_c$ such that $g_c(r_{\mathrm{row}})$ matches a
fabricated leaf-claim.  Because $\gamma$ is sampled after
$r_{\mathrm{row}}$ and after the per-interaction projections are
fixed, the prover has no degrees of freedom: the malicious
$g_c(r_{\mathrm{row}})$ would have to equal a specific value over
$\gamma$-randomness, which holds with probability $1/|\mathbb{F}|$
per column.  Across $w_c$ columns, soundness error is
$w_c / |\mathbb{F}|$ --- negligible for $|\mathbb{F}| > 2^{124}$.

\subsection{The VEIL Multilinear PCS Construction (the actual answer)}

After the dead end above, a survey of the SP1 VEIL paper
\cite{veil2026} (Dalal--Hemo--Rabinovich--Rothblum,
\texttt{slop/crates/veil/paper/veil.pdf}) revealed Lemma~4.7 (ZK
Multilinear PCS), which is exactly the late-binding multi-column
construction we need.  The ZK aspect is incidental --- the underlying
late-binding wrapper composes with any Reed--Solomon-based base PCS,
including WHIR, and remains sound for the non-ZK case (we drop the
masking padding $k=0$).

\paragraph{Construction overview.}
Let $f \in \mathbb{F}^{2^n}$ be the evaluation vector of the
$n$-variable multilinear polynomial we want to commit (e.g.~the chip
trace serialised in row-major order).  Pick a stacking parameter
$0 < p < n$.

\paragraph{Commit phase.}
\begin{enumerate}
  \item Decompose $f$ into $2^p$ stacks $f_\ell$ of length
        $2^{n-p}$ ($f_\ell$ is the slice where the leading $p$
        coordinates equal $\ell$).  In our STARK use case
        $2^p$ = number of chip columns and $2^{n-p}$ = trace height.
  \item (Optional ZK pad: append $k$ random elements to each $f_\ell$
        and a fresh row $f_{2^p+1}$.  We use $k=0$.)
  \item Encode each stack: $C f_\ell$ via the base RS code.
  \item Send the interleaved codeword
        $\mathfrak{C}_{\ell,i} = (C f_\ell)_i$ as the merkle commit.
\end{enumerate}

\textbf{Crucially, no $\alpha$-batching challenge is sampled at
commit time, and no opening points are absorbed yet.}  The merkle
commit is purely structural over the per-column codewords.

\paragraph{Open at a post-commit point $x = (x^{(1)}, x^{(2)})$.}
$x^{(1)} \in \mathbb{F}^p$ is the column-index dimension,
$x^{(2)} \in \mathbb{F}^{n-p}$ is the row dimension (in our case
$x^{(2)} = r_{\mathrm{row}}$ from GKR).
\begin{enumerate}
  \item Prover exposes per-stack values
        $y_\ell := f_\ell(x^{(2)})$ for $\ell \in \{0,1\}^p$.  These
        are the per-column row-MLE values --- exactly what
        \texttt{LogUpRowOpening} carries.
  \item Verifier samples $\vec{\rho} \leftarrow \mathsf{PG}$
        \emph{after} seeing the exposed $y_\ell$.  This is the binding
        step: $\vec{\rho}$ is the column-batching randomness, sampled
        post-late-point and post-exposure.
  \item Prover sends explicit $g'$, the linear combination
        $g'(\cdot) := \sum_\ell \rho_\ell f_\ell(\cdot)$ over the
        coordinate dimension.
  \item Verifier defines a virtual oracle
        $\mathfrak{C}_\rho$ via $\mathfrak{C}_{\rho,i}
        = \sum_\ell \rho_\ell \mathfrak{C}_{\ell,i}$, which can be
        resolved by querying the merkle-committed $\mathfrak{C}$.
  \item Run the base evaluation PCS (\textbf{this is where WHIR
        slots in}) to prove $g(x^{(2)}) = a$ where $a := \sum_\ell
        \rho_\ell y_\ell$ is the verifier-recomputable target.
  \item Verifier checks:
        \begin{enumerate}[(a)]
          \item $\hat{Y}(x^{(1)}) = y$ where $\hat{Y}$ is the
                multilinear extension of
                $Y: \{0,1\}^p \to \mathbb{F}: \ell \mapsto y_\ell$
                and $y$ is the user-claimed evaluation.  This binds
                the per-column values to the requested point in the
                column dimension.
          \item $\sum_\ell \rho_\ell y_\ell = a$ (the explicit linear
                check on exposed values).
        \end{enumerate}
\end{enumerate}

\paragraph{Why this works (and the previous design didn't).}
The fatal gap in the naive multi-column late-binding was that
$\alpha$ (column-batching) had to be sampled at commit time, before
the late opening point was known.  A malicious prover, knowing
$\alpha$ and the eventual late point $r_{\mathrm{row}}$, could pick
per-column values $v_j$ to satisfy $\sum_j \alpha^j v_j =
g(r_{\mathrm{row}})$, gaining $w-1$ degrees of freedom.

VEIL Lemma~4.7 inverts the order:
\begin{enumerate}
  \item Commit \emph{without} any batching challenge.
  \item Late-derive $r_{\mathrm{row}}$.
  \item Prover \emph{exposes} per-column values $y_\ell$ first.
  \item Verifier samples $\vec{\rho}$ \emph{after} seeing both
        $r_{\mathrm{row}}$ \emph{and} the $y_\ell$.
  \item The check $\sum_\ell \rho_\ell y_\ell = a$, where $a$ is
        bound by the base PCS opening of $g(r_{\mathrm{row}})$, makes
        any single-column lie detectable with probability
        $1 - 1/|\mathbb{F}|$ over $\vec{\rho}$.
\end{enumerate}

The verifier can compute $a$ from $\vec{\rho}$ and the exposed $y_\ell$;
the base PCS proof binds $g(r_{\mathrm{row}}) = a$ to the merkle
commitment via the virtual oracle $\mathfrak{C}_\rho$.  All per-column
values are bound.

\paragraph{Cost analysis (Lemma 4.7, $k=0$).}
\begin{itemize}
  \item Commit time: $(2^p + 1) \cdot T_C(2^{n-p})$ --- one RS encode
        per column.
  \item Prove time: $T_{P_0}(2^{n-p}) + O(2^n)$ --- base WHIR proof
        on a single $2^{n-p}$-size polynomial plus the linear
        combination.
  \item Verify time: $T_{V_0}(2^{n-p}) + O(k_{0,n-p} \cdot 2^p)$.
  \item Message complexity: $2^p + 2$ field elements (the exposed
        $y_\ell$ values plus base PCS messages).
\end{itemize}

For a typical chip with $w \approx 32$ columns and $h \approx 2^{17}$
rows ($n=22$, $p=5$, $n-p=17$):
\begin{itemize}
  \item Commit: $33$ RS encodes of $2^{17}$ values --- comparable to
        the existing batched FRI commit (which encodes the full
        $32 \times 2^{17}$ matrix as a single codeword anyway).
  \item Prove: one WHIR proof of size $\sim$few~KB plus $\sim 32$
        extra field elements.
  \item Verify: standard WHIR verify cost plus $\sim 32$
        coefficient operations.
\end{itemize}

The overhead vs.~plain WHIR is the $2^p \approx 32$ exposed values
plus a small linear-algebra check --- negligible compared to the
$\sim 30\times$ cost of per-column commits (option A) or the
research-grade WHIR variant required for option B.

\paragraph{Mapping to our STARK pipeline.}
\begin{itemize}
  \item Per chip, the main trace matrix is $w$ columns by $h$ rows,
        i.e.~an evaluation vector of size $w \cdot h$ over
        $n = \log_2(w) + \log_2(h)$ variables.  Set $p = \log_2 w$.
  \item Commit via the interleaved codeword (replaces the existing
        WHIR fast-path \texttt{pcs.commit}).
  \item After zerocheck and LogUp-GKR derive $r_{\mathrm{row}}$, run
        the open-at-point protocol with $x^{(2)} = r_{\mathrm{row}}$
        and $x^{(1)}$ = the $p$ coordinates that select the column
        index.  The exposed $y_\ell$ values are exactly
        \texttt{LogUpRowOpening.main\_at\_r\_row}.
  \item The verifier reconstructs the leaf claim from the
        bound $y_\ell$ exactly as the existing
        \texttt{reconstruct\_leaf\_claim\_from\_openings} does ---
        no GKR restructuring needed.
\end{itemize}

\subsection{Cross-Binding Plan (Closing the First-Iteration Gap)}
\label{sec:cross-binding-plan}

The current shipped prover/verifier wiring uses a fresh challenger
for the late-binding step, leaving the WHIR commit independent of
the FRI commit at $\zeta$.  Concrete attack: a malicious prover
commits trace $A$ via FRI (used by zerocheck and the existing
\texttt{pcs.verify}) and trace $B \neq A$ via WHIR late-binding
(used to bind \texttt{LogUpRowOpening.main\_at\_r\_row}).  The
verifier accepts both because:
\begin{itemize}
  \item zerocheck checks $A$'s transition constraints on its
        hypercube;
  \item GKR + late-binding check $B$'s lookups balance.
\end{itemize}
A program execution requires \emph{both} for the same trace; the
gap lets the prover decouple them.

\paragraph{Two viable closures (in increasing scope).}

\textbf{(C.1) Brakedown-style spot checks (smaller change).}
After both commits are made, sample $K$ random trace positions
$i_1, \ldots, i_K \in \{0, \ldots, 2^h - 1\}$.  For each position
$i_j$:
\begin{itemize}
  \item Prover provides $\mathrm{trace}[i_j]$ (a length-$w$ vector
        of per-column values).
  \item Prover provides merkle paths against the FRI commit's
        codeword leaves at the codeword position(s) corresponding to
        $i_j$ (the FRI commit is on the LDE codeword; recovering
        $\mathrm{trace}[i_j]$ from the codeword requires the
        evaluation domain to include $i_j$ as a coset point, which
        it does for the natural-domain coset).
  \item Prover provides merkle paths against each per-column WHIR
        commit at the codeword position corresponding to $i_j$.
\end{itemize}
Verifier checks: both sides give the same $\mathrm{trace}[i_j]$ for
each $i_j$.  Soundness: if $A \neq B$ they differ at $\geq 1$ trace
position; $K$ random positions catch the divergence with probability
$1 - (1 - 1/N)^K$.  For $K = 80$, $N = 2^{17}$, the missed-detection
probability is $\sim 2^{-80}$.

Cost: $K \cdot \log_2(N \cdot \text{rate})$ extra hashes per side
per chip.  For typical chips (\(K = 80\), \(N = 2^{17}\),
\(\text{rate} = 16\)): $\sim 1700$ hashes per chip per side, so
$\sim 100\,$kB extra per chip per shard.  Manageable.

\paragraph{C.1 implementation reality check.}
The Brakedown-style spot-check sounds simple but hits a real impedance
mismatch: the FRI commit's MMCS opens at LDE-codeword positions on a
multiplicative-subgroup coset (rate 2 typically); the WHIR commit's
MMCS opens at RS-codeword positions on a different coset (rate 16
in our config).  ``Trace position $i$'' is one merkle path in each,
but the literal merkle leaves are codeword values that interpolate
to $\mathrm{trace}[i]$, not $\mathrm{trace}[i]$ itself.  Direct
spot-check at canonical trace positions requires:
\begin{itemize}
  \item Mapping trace[i] to (FRI codeword position, WHIR codeword
        position) via the respective coset rotation / RS encoding.
  \item Querying both MMCS at those positions and de-rotating.
  \item Re-checking interpolation consistency.
\end{itemize}
This is implementable but requires touching internal PCS plumbing in
both \texttt{p3-fri} and \texttt{p3-whir}.

\paragraph{C.1 WHIR-side: shipped.}
\begin{itemize}
  \item \texttt{CrossBindingSpotCheckProof} wire format
  \item \texttt{prove\_cross\_binding\_spot\_check(dense, whir\_state, ch, K)}
        opens the WHIR commit's MMCS at K spot-check positions via
        \texttt{Mmcs::open\_batch} and serialises the
        \texttt{BatchOpening} per position.
  \item \texttt{verify\_cross\_binding\_spot\_check\_whir(proof, commit, height, width)}
        runs \texttt{Mmcs::verify\_batch} per position and returns
        true iff every WHIR-side merkle path validates against the
        parsed WHIR commitment.
  \item Cryptographic tamper rejection: mutating one merkle path byte
        causes \texttt{verify\_cross\_binding\_spot\_check\_whir} to
        return false (test \texttt{cross\_binding\_whir\_rejects\_tampered\_path}).
\end{itemize}

\paragraph{C.1 FRI-side: not yet started.}
The FRI commit's MMCS lives inside \texttt{StarkConfig::Pcs}'s
prover data (\texttt{ShardMainData::main\_data}); accessing it for
spot-check generation requires either:
\begin{itemize}
  \item Exposing the underlying MMCS from \texttt{p3-fri}'s
        \texttt{TwoAdicFriPcs}, OR
  \item Threading the MMCS through the StarkConfig generic chain.
\end{itemize}
Both are several-hour PRs on the FRI side.  Status: deferred pending
the architectural choice between completing C.1 (FRI plumbing) and
landing C.2 (drop FRI in WHIR fast path; WHIR commit becomes sole
binding).

\textbf{(C.2) Single-commit migration (cleaner, larger change).}
Drop the FRI commit entirely from the WHIR fast path.  The WHIR
late-binding commit becomes the sole binding for the main trace.
Required changes:
\begin{itemize}
  \item In \texttt{prover.rs::open}, skip \texttt{pcs.commit} +
        \texttt{pcs.open} for the main trace; set
        \texttt{opening\_proof} to a placeholder, set
        \texttt{ShardOpenedValues.chips[*].main} to empty.
  \item In \texttt{verifier.rs::verify\_shard}, skip
        \texttt{config.pcs().verify(...)} when WHIR fast path is in
        use; rely entirely on the late-binding verification for
        main-trace soundness.
  \item Add a second late-binding opening at \emph{zerocheck}'s
        $r_{\mathrm{zerocheck}}$ to bind the zerocheck final claim
        to the trace (currently the zerocheck final claim is
        transcript-bound but not PCS-bound; opens an analogous gap
        to the GKR one).
\end{itemize}
Soundness: clean (one commit, all openings against it).  Cost:
no extra commit (saves the FRI commit + open time).  Risk: changes
the on-wire \texttt{ShardProof} shape and the recursion-circuit
verifier.

\paragraph{Recommendation.}
Ship (C.1) first as a bounded incremental fix that closes the
soundness gap with the current architecture; (C.2) is the long-term
target once the rest of the WHIR fast path stabilises and the
recursion circuit can be updated alongside.

\subsection{Implementation Plan}

\begin{enumerate}
  \item \textbf{Single-column late-binding adapter (DONE).}
        \texttt{crates/stark/src/whir\_late\_binding.rs}, two passing
        tests --- foundation for the multi-column wrapper.
  \item \textbf{Per-column multi-column adapter (DONE, simpler than
        VEIL).}
        \texttt{commit\_multi\_column} /
        \texttt{add\_late\_claim\_multi\_column} /
        \texttt{open\_multi\_column} loop the single-column adapter
        with a shared challenger.  Soundness clean (each column's
        WHIR commitment binds its own row-MLE value).  Cost: $w$ WHIR
        proofs per chip vs.~one batched proof from upstream.  Caller
        API matches the eventual interleaved-codeword optimisation.
        Two new tests (\texttt{late\_binding\_round\_trip\_multi\_column}
        and \texttt{late\_binding\_multi\_column\_rejects\_tampered\_value}),
        both passing.
  \item \textbf{Convenience helpers (DONE).}
        \texttt{prove\_chip\_late\_binding} and
        \texttt{verify\_chip\_late\_binding} wrap the per-column
        adapter at fixed KoalaBear types
        (\texttt{ZirenLateBindingPcs} alias).  The \texttt{derive\_r\_row}
        closure lets the caller plug in the GKR-derived row
        coordinates without coupling the adapter to GKR.
  \item \textbf{Serialisation helpers (DONE).}
        \texttt{serialize\_chip\_proofs} /
        \texttt{deserialize\_chip\_proofs} use \texttt{serde\_json}
        because the WhirProof's \texttt{QueryOpening} enum is
        internally tagged (\texttt{tag = "type"}), which bincode 1.x
        cannot deserialize.  JSON is verbose but portable; production
        should switch to a binary self-describing format
        (\texttt{rmp-serde} / msgpack) once added as a workspace dep.
        Round-trip test
        \texttt{late\_binding\_serialization\_round\_trip} confirms
        binding holds across the JSON wire format.
  \item \textbf{Trait dispatch (DONE).}
        \texttt{LateBindingCapable} extends
        \texttt{StarkGenericConfig} with two methods:
        \texttt{prove\_chip\_late\_binding} and
        \texttt{verify\_chip\_late\_binding}.
        \texttt{KoalaBearPoseidon2Inner} implements the trait by
        delegating to the convenience helpers.  This is the
        architectural bridge that lets the generic
        \texttt{prover.rs} call WHIR-specific code without exposing
        WHIR types in \texttt{StarkGenericConfig} and without
        unsafe transmute.
  \item \textbf{Full wiring simulation test (DONE).}
        \texttt{late\_binding\_full\_wiring\_simulation} mirrors what
        \texttt{prover.rs} will do: 3 chips with varying
        (height, width), per-chip late-binding with a shared
        Fiat-Shamir challenger, verifier-side replay; validates the
        challenger advances correctly across chips and the
        prover/verifier derive the same $r_{\mathrm{row}}$.
  \item \textbf{ShardProof field scaffolding (DONE).}
        \texttt{ShardProof::late\_binding\_proofs:
        Option<Vec<Vec<u8>>>} added with
        \texttt{\#[serde(default)]} for backward-compat.  All
        existing call sites set the field to \texttt{None}; no
        behaviour change yet.
  \item \textbf{Interleaved-codeword multi-column adapter (TODO,
        optimisation).}
        Following VEIL Lemma~4.7: commit each chip's main trace as
        an interleaved codeword with no batching at commit, then
        late-bind via the post-commit $\vec{\rho}$ protocol.  Reduces
        per-chip proof from $w$ WHIR proofs to 1 base WHIR proof +
        $w$ exposed field elements.  No API change vs.~the per-column
        adapter.
  \item \textbf{Prover wiring (TODO).}
        In \texttt{prover.rs}, after the LogUp-GKR step computes
        \texttt{logup\_row\_openings}, dispatch through the
        \texttt{LateBindingCapable} trait
        (\texttt{<SC as LateBindingCapable>::prove\_chip\_late\_binding(\ldots)})
        to produce the per-chip serialised bytes and store them in
        \texttt{ShardProof::late\_binding\_proofs}.  The
        \texttt{derive\_r\_row} closure pulls
        \texttt{logup\_gkr\_proofs[i].eval\_point[..num\_vars]}.
        Required: a way to express the trait bound on
        \texttt{prove\_shard} without breaking call sites that use
        non-LateBindingCapable SCs --- either a separate
        \texttt{prove\_shard\_with\_late\_binding} entry point, or a
        new trait that all SCs implement (with a default no-op for
        non-WHIR configs).
  \item \textbf{Verifier wiring (TODO).}
        In \texttt{verifier.rs}, after the LogUp-GKR replay,
        dispatch through
        \texttt{<SC as LateBindingCapable>::verify\_chip\_late\_binding(\ldots)}
        with each chip's bytes, the row-MLE values from
        \texttt{logup\_row\_openings.main\_at\_r\_row}, and the same
        \texttt{derive\_r\_row} closure shape.  The existing
        \texttt{reconstruct\_leaf\_claim\_from\_openings} stays
        unchanged because the per-column row-MLE values it consumes
        are now cryptographically bound by the trait dispatch.
  \item \textbf{Recursion circuit (deferred).}
        Update \texttt{verify\_whir\_pcs} in
        \texttt{crates/recursion/circuit/src/whir\_verifier.rs} to
        consume the new opening proof shape (interleaved-codeword
        merkle root + exposed $y_\ell$ + post-commit $\vec{\rho}$
        protocol + base WHIR proof).
\end{enumerate}
